---
title: "HW2-EDA with R - Part1: Plotting"
author: "misken"
date: "January 25, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.

## Step 3 - Set working directory and explore data folder

Set the working directory or create an R Studio project in the current folder. You'll notice that there is a folder named **data**.
Inside of it you'll find the data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data. **You do NOT need to rerun the census_data_prep.R file.** I'm
including it and encouraging you to take a look through it to get a sense of what a
typical data prep script might look like. We'll be learning to do all of the things
shown in that file.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.

Note that if you have errors in your R code, **knitr** will stop at the error
and won't create an html file. Sooo, you should comment out R code that has
errors and write a little comment stating that it's an error. Keep doing this
until you successfully knit an html file based on your R Markdown document.
You'll be submitting both the html file as well as the Rmd file.

You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(dplyr)
library(ggplot2)
library(ggthemes) 
# Make your ggplots look even better
```


### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_MI.csv** into a data frame named county_MI.

```{r read_countydata}
county_df <- read.csv("data/county.csv", header = TRUE, stringsAsFactors = TRUE)
county_MI <- read.csv("data/county_MI.csv", header = TRUE, stringsAsFactors = TRUE)
```

* What does each row represent?

> Each row represents the observation (data) for the county that is listed in "County" column.

Check the structure of each data.frame and use the `summary` function to begin
exploring the data. 

* Are there any NA values? 

> In "county_df", we have NAs in *Income*, *IncomeErr*, and *ChildPoverty* data variables.There is one NA value in each one of these three data sets.
  In "county_MI", we do not have any NAs.

* What differences do you notice between the structure of the two data frames?

> With the restriction to one state (MI), *"county_MI"* has less observations. However, we have an additional variable that is the *"peninsula"*. Therefore, the two data frames are not the same size. The "peninsula" variable groups the counties in MI to represent the upper and lower peninsulas.


```{r}

str(county_df)
summary(county_df)

str(county_MI)
summary(county_MI)

```

The county_MI data frame contains data for Michigan only and includes an
peninsula field. See **census_data_prep.R** to see how this field was
created. It involves scraping a table out of a web page.

**Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.**

### Problem 2 - Basic plots for median Income

Note that the field called `Income` is the median income
for that county. You do **NOT** need to compute a median. The values are the medians.

Create a histogram, boxplot, kernel density plot and violin plot for median
income for the entire US. 
Write a short paragraph describing the distribution of median incomes.

> We have a right skewed histogram that typically mean is less than the median, however, in this case mean (46,830) is slightly bigger than the median (45,111). This can be the result of very high median incomes that show up in the tail. We see the peak between 37,500 to 50,000 dollars that aligns with the 2nd and 3rd quartiles' shape in the violin plot.

```{r}
# ggplot(???) + geom_histogram(???)

#binwidth calculation below is used for histogram plot

b_width <- round((max(county_df$Income, na.rm = TRUE) - min(county_df$Income, na.rm = TRUE)) / (sqrt(length(county_df$Income))))

g_cdf <- ggplot(data = county_df)

g_cdf + 
  geom_histogram(aes(x= Income), na.rm = TRUE, binwidth = b_width, fill = "darkblue", colour = "darkblue") + 
  ggtitle("Histogram of Median Income")

g_cdf + 
  geom_boxplot(aes(x= " ", y= Income), na.rm = TRUE, fill = "darkblue", colour = "black") + 
  xlab("US Counties") + 
  ggtitle("Boxplot of Median Income")

ggplot(county_df, aes(x= Income)) + 
  geom_density(na.rm = TRUE, colour= "darkblue") + 
  ggtitle("Density plot of Median Income")

ggplot(county_df, aes(x= " ", y= Income)) + 
  geom_violin(na.rm = TRUE, fill= "darkblue") + 
  xlab("US Counties") + 
  ggtitle("Violin plot of Median Income")

```

### Problem 3 - Histogram of total population

Create a histogram of total population for counties in the US. 

```{r histo_totpop}
#bin calculation below is used for histogram plot

bin_var <- round((sqrt(length(county_df$Income))))

g_cdf + 
  geom_histogram(aes(x= TotalPop), bins = bin_var, fill = "darkblue", colour = "darkblue")

```

Repeat the histogram about but make the following style changes to the plot.

* Add a title such as "Distribution of population for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Total Population"
* Change the colors of the bars to something other than the default black

```{r histo_stylechanges}

theme_update(plot.title= element_text(hjust= 0.5))

g_cdf + 
  geom_histogram(aes(x= TotalPop), bins= bin_var, fill = "darkblue", colour = "darkblue") + 
  ggtitle("Distribution of population for US counties") + 
  ylab("Number of counties") + 
  xlab("Total Population")

```

Repeat the histogram but only for counties with less than 1.5 million people. Do NOT use
dplyr to do the filtering - instead, use boolean indexing (see intro to dataframes Rmd file).

```{r histo_lt1.5m}

ggplot(data = county_df[county_df$TotalPop < 1500000, ])+ 
  geom_histogram(aes(x= TotalPop), bins= bin_var, fill = "darkblue", colour = "darkblue") + 
  ggtitle("Distribution of population for US counties with population < 1.5 million") + 
  ylab("Number of counties") + 
  xlab("Total Population")

```


**Hacker Extra** - Create a function that uses the
[Freedman-Diaconis](https://en.wikipedia.org/wiki/Freedman%E2%80%93Diaconis_rule) rule for calculating bin width and number of bins. The function should take a vector (e.g. a dataframe column) as its
input and return the integer number of bins recommended for the histogram.

```{r}
numbins_FD <- function(data_col){
  
 (num_bins <- round((max(data_col) - min(data_col)) / (2*IQR(data_col)*(length(data_col)^(-1/3)))))
  
}

numbins_FD(county_df$TotalPop)

nclass.FD(county_df$TotalPop)

```



### Problem 4 - How do Michigan's peninsulas compare?

Create a grouped violin plot showing Income by peninsula. Make sure
your graph has a title and good axis labels.

Do the peninsulas 
differ with respect to distribution of median income over the counties?

> The distribution of median income differs between the peninsulas of MI. *Lower peninsula* shows more variance with its wider range from 30,000 to 75,000 and its spread seen via the shape of the violin. On the other hand, *upper peninsula* shows a condensed grouping of median income level that is approximately between 34,000 and 46,000.


```{r vp_inc_by_peninsula}

ggplot(county_MI, aes(x= peninsula, y= Income)) + 
  geom_violin(na.rm = TRUE, fill= "darkblue") + 
  ggtitle("Distribution of Income grouped by MI peninsulas") + 
  xlab("MI peninsulas") + 
  ylab("Income")

```

Create several boxplots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

> **Population by peninsulas**

> **Commute time by peninsulas**

> **Production by peninsulas**

> **Poverty by peninsulas**


```{r boxplots}

ggplot(county_MI, aes(x= peninsula, y= TotalPop)) + 
  geom_boxplot(na.rm = TRUE, fill= "darkblue", colour = "black") + 
  ggtitle("Distribution of Population grouped by MI peninsulas") + 
  xlab("MI peninsulas") + 
  ylab("Total Population")


ggplot(county_MI, aes(x= peninsula, y= MeanCommute)) + 
  geom_boxplot(na.rm = TRUE, fill= "darkblue", colour = "black") + 
  ggtitle("Distribution of Commute by MI peninsulas") + 
  xlab("MI peninsulas") + 
  ylab("Mean Commute Time")

ggplot(county_MI, aes(x= peninsula, y= Production)) + 
  geom_boxplot(na.rm = TRUE, fill= "darkblue", colour = "black") + 
  ggtitle("Distribution of % population employed in Production or Transportation") + 
  xlab("MI peninsulas") + 
  ylab("% Pop in Production or Transportation")

ggplot(county_MI, aes(x= peninsula, y= Poverty)) + 
  geom_boxplot(na.rm = TRUE, fill= "darkblue", colour = "black") + 
  ggtitle("Distribution of % of population under poverty level by MI peninsulas") + 
  xlab("MI peninsulas") + 
  ylab("% of population under poverty level")

```

In county_df you'll see that there's a Region field. Create a set of faceted 
histograms showing how unemployment rate compares by region.  Make sure your graph has a title and good axis
labels. 

```{r faceted_histo}
ggplot(data = county_df) + 
  geom_histogram(aes(x= Unemployment), bins= 56, fill = "darkblue", colour = "darkblue") + 
  ggtitle("Distribution of Unemployment Rate by US Regions") + 
  ylab(" ") + 
  xlab("Unemployment Rate") + 
  facet_grid(~ Region)

```

Now do a group of boxplots of mean commute time by region.

```{r bp_commute_region}
ggplot(data = county_df) + 
  geom_boxplot(aes(x= "", y= MeanCommute), fill = "darkblue", colour = "black") + 
  ggtitle("Distribution of Mean Commute Time by US Regions") + 
  xlab("") + 
  ylab("Mean Commute Time") + 
  facet_grid(~ Region)
  
```

### Problem 6 - Basic bar chart

Create a bar chart showing the number of counties by Division in the US. Make sure
your graph has a title and good axis labels. Figure out how to format the
x-axis labels so that they don't obscure each other.

```{r bar_counties_by_division}

ggplot(county_df, aes(Division)) + 
  geom_bar(stat = 'count', colour = "darkblue") + 
  ggtitle("Number of Counties by Division in the US") + 
  ylab("# of Counties") + 
  xlab("") + 
  theme(axis.text.x=element_text(angle=90, vjust=1, hjust=0))

```

### Problem 7 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
Income and MeanCommute.

```{r xy_inc_commute}

ggplot(county_MI, aes(Income, MeanCommute)) + 
  geom_point(color = "darkblue") + 
  ggtitle("Income relation to Commute Time (in MI)") + 
  xlab("Income") + 
  ylab("Commute Time")
    
```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula and the size is mapped to total population.

```{r xy_inc_commute_peninsula}

ggplot(county_MI, aes(Income, MeanCommute, color= peninsula, size= TotalPop)) + 
  geom_point() + 
  ggtitle("Income relation to Commute Time by Peninsula (in MI)") + 
  xlab("Income") + 
  ylab("Commute Time") + 
  geom_smooth(method = 'lm')

```

Your thoughts on the relationship between these variables?

> Upper peninsula points are more condensed in the lower left side of the plot and do not indicate a relation between income level and commute time. The size of the Upper peninsula points, which are tied to total population, also do not show much variation at different income levels or commute time.

> Lower peninsula data displays a relation between income levels and commute time. We see the points spreading to the upper part of the plot as the income levels increase. For the relation between commute time and population, we see that the largest points are at or above 25 minutes. That aligns with the expectation to see higher commute times in crowded or highly populated areas.

Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?

```{r xy_faceted_peninsula Income and Production}

ggplot(county_MI, aes(Income, Production, color= peninsula)) + 
  geom_point() + 
  ggtitle("Income relation to production sector employment by MI Peninsulas") + 
  xlab("Income") + 
  ylab("Production emp") + 
  geom_smooth(method = 'lm') + 
  facet_grid(~peninsula)

```




```{r xy_faceted_peninsula Total population and Asian pop}

ggplot(county_MI, aes(log(TotalPop), log(Asian), color= peninsula)) + 
  geom_point() + 
  ggtitle("Total Population relation to Asian population by MI Peninsulas") + 
  xlab("Total Population (log)") + 
  ylab("Asian (log)") + geom_smooth(method = 'lm', na.rm = TRUE) + 
  facet_wrap(~peninsula)

```


```{r xy_faceted_peninsula Women population and Drive}

ggplot(county_MI, aes(log(Women), log(Drive), color= peninsula)) + 
  geom_point() + 
  ggtitle("Women Population relation to driving average by MI Peninsulas") + 
  xlab("Women Population (log)") + 
  ylab("Drive Avg (log)") + 
  geom_smooth(method = 'lm', na.rm = TRUE) + 
  facet_wrap(~peninsula)

```


Now use the full US data frame.

Create a scatter plot of TotalPop by Income.

```{r xy_pop_inc}

ggplot(county_df, aes(Income, TotalPop)) + 
  geom_point(na.rm = TRUE, color= "darkblue") + 
  ggtitle("Income relation to Population in US") + 
  xlab("Income") + 
  ylab("Total Population")

```

The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.


```{r xy_pop_inc_log}

ggplot(county_df, aes(log(Income), log(TotalPop))) + 
  geom_point(na.rm = TRUE, color= "darkblue") + 
  ggtitle("Income relation to Population in US") + 
  xlab("Income (log)") + 
  ylab("Total Population (log)") + 
  geom_smooth(method = 'lm', na.rm = TRUE)

```



### Hacker Extra - More analysis

Do some additional plotting to look for interesting patterns or relationships
either in the full US dataset or the MI only dataset. In addition, try out
some of ggplots additional features such as themes, axis formatting, gridline
control, or others.